### 2020-08-31 thesis/chap_2.tex
small fixes

This research is an extension of the work intitiated by the project “Montérégie Connection”, whose goal was to engage stakeholders around a discussion of the ecosystem services provided by natural habitat in the region. In the context of this stakeholder-driven project, \cite{mitchell_monteregie_2015} conducted a non-spatially explicit scenario building exercise for a subset of the Montérégie region, the Vallée-de-Richelieu Municipalité Region Comté. The outcome was 4 scenarios, ranging from “Periurban Development” to “Green Development”, and each describing different land use change trends. The Montérégie connection project improved our understanding of the relationship between forest connectivity and ecosystem services. However, there was no effort to estimate the impact of each scenario on ecological connectivity. This is partially due to a lack of tools able to translate these scenarios into simulations. We contribute to the development of such tools by integrating the results of the workshop into the land use change model developed in chapter 1. The goal of this chapter is therefore twofold: gain a better understanding of conservation priorities in the region, and to provide a basis for an iterative scenario-building process. \\
%-----------------------------------------------

@@

In order to understand what stakeholders perceive as a conservation priority, and derive conservation scenarios from those perceptions, we conducted a community workshop. The workshop was approved by McGill University’s Research Ethics Board (see appendix {x} for the approval letter). The workshop was held on January 22nd 2020 in the public library of Saint-Jean sur Richelieu. \\

@@

Each participant received a unique and neutral identifier of the form Affiliation-Geography. For example, a (hypothetical) land use planner from the table that brought together participants from the Maskoutains region received the code [A-1-1], and the UPA representatives from the Richelieu region received the codes [C-3 -1] and [C-3-2]. In addition, for certain activities, the affiliation of the participants is color-coded. For instance, blue for the land use planners and purple for the ministry representatives.

@@

In order to integrate the results with the land use change model, we added a 5 km buffer around the linkages to simulate a corridor effect. This corridor width size is arbitrary and is in most cases  an unrealistic expectation forhow large a wildlife corridor can or should be. \\

@@

The reforestation status modified what zones can be targeted for reforestation. In reforestation scenarios in Chapter 1, reforestation happened randomly (barring some neighboring rules). In this chapter, reforestation is targeted to the same areas identified in the workshop (all pixels within the identified priority areas and within a 5km buffer around the identified priority linkage areas). As in the model described in Chapter 1, all new reforested area takes the value of medium aged (30-50 yr) deciduous forest and remain in this state for the rest of the simulation. As previously mentioned, this assumption is necessary because LANDIS surface types are not available for newly forested areas. For the same reason, no forest transitions are allowed for this forest type. These two assumptions are conservative:  we do not impose succession or disturbance on these newly forested ares and assume medium age as an average age class.

@@

In order to be able to compare the results of these new scenarios with those developed in Chapter 1, we conducted the same connectivity analyses using the Circuitscape software. The software “borrows algorithms from electronic circuit theory to predict connectivity in heterogeneous landscapes''. It is a free and open software under MIT license developed originally in Python, and now in its 5th version, in Julia \citep{circuitjulia}.

@@

Participants also workshopped constraints and assets to connectivity conservation in Montérégie. We divided the results into elements that had been spatialized (i.e. placed on a specific location of the map during the exercise), and non spatialized (i.e. not placed on a specific location ), and scored them for the level of consensus they generated. Constraints scored higher, on average, than assets, indicating a focus of participants on constraints. The elements in the spatialized set scored higher on average than in the non spatialized set, indicating that participants tended to focus on more tangible and local issues and assets.

@@

Participants debated the importance of both "facilitating" and "constraining" factors in deciding whether a link is a priority. Facilitating factors included the feasibility of establishing the link, social facilitators such as political and social buy-in, and territorial facilitators, such as appropriate land use. There were several types of constraints: financial, social, territorial and physical, and must be considered in relation to the socio-political and spatial planning context in which they emerge. % example?

@@

% Linear
\begin{figure}[h!]
\makebox[\textwidth]{
  \includegraphics[width=1.3\textwidth]{thesis/figures/connectivity_decrease_x5species_chap2.png}
}
 \caption{Change in mean flow (in \% of the the 2010 flow) between 2010 (observed) and 2100 (predicted), contrasting BAU scenario (solid line) with other conservation scenarios - linear graph.}
 \label{fig:flow_linear_2}
\end{figure}

% Radar
\begin{figure}[h!]
\makebox[\textwidth]{
  \includegraphics[width=1.3\textwidth]{thesis/figures/radar_ggradar_chap2.png}
}
 \caption{Change in mean flow (in \% of the the 2010 flow) between 2010 (observed) and 2100 (predicted), contrasting BAU scenario (solid line) with other conservation scenarios - radar graph.}
 \label{fig:flow_radar_2}
\end{figure}

@@

% Linear
\begin{figure}[h!]
\makebox[\textwidth]{
  \includegraphics[width=1.3\textwidth]{thesis/figures/surf_chap2.png}
}
 \caption{Change in the number of features (in \% of the the 2010 flow) identified by the SURF analyssis between 2010 (observed) and 2100 (predicted), contrasting BAU scenario with other conservation scenarios - linear graph.}
 \label{fig:surf_linear_2}
\end{figure}

% Radar
\begin{figure}[h!]
\makebox[\textwidth]{
  \includegraphics[width=1.3\textwidth]{thesis/figures/surf_radar_chap2.png}
}
 \caption{Change in the number of features (in \% of the the 2010 flow) identified by the SURF analyssis between 2010 (observed) and 2100 (predicted), contrasting BAU scenario with other conservation scenarios - radar graph.}
 \label{fig:surf_radar_2}
\end{figure}

>>>>

This research is an extension of the work initiated by the project “Montérégie Connection”, whose goal was to engage stakeholders around a discussion of the ecosystem services provided by natural habitat in the region. In the context of this stakeholder-driven project, \cite{mitchell_monteregie_2015} conducted a non-spatially explicit scenario building exercise for a subset of the Montérégie region, the Vallée-de-Richelieu Municipalité Region Comté. The outcome was 4 scenarios, ranging from “Periurban Development” to “Green Development”, and each describing different land use change trends. The Montérégie connection project improved our understanding of the relationship between forest connectivity and ecosystem services. However, there was no effort to estimate the impact of each scenario on ecological connectivity. This is partially due to a lack of tools able to translate these scenarios into simulations. We contribute to the development of such tools by integrating the results of the workshop into the land use change model developed in chapter 1. The goal of this chapter is therefore twofold: gain a better understanding of conservation priorities in the region, and to provide a basis for an iterative scenario-building process. \\
%-----------------------------------------------

@@

In order to understand what stakeholders perceive as a conservation priority, and derive conservation scenarios from those perceptions, we conducted a community workshop. The workshop was approved by McGill University’s Research Ethics Board (see Appendix 2 for the approval letter). The workshop was held on January 22nd 2020 in the public library of Saint-Jean sur Richelieu. \\

@@

Each participant received a unique and neutral identifier of the form Affiliation-Geography. For example, a (hypothetical) land use planner from the table that brought together participants from the Maskoutains region received the code [A-1-1], and the UPA representatives from the Richelieu region received the codes [C-3-1] and [C-3-2]. In addition, for certain activities, the affiliation of the participants is color-coded. For instance, blue for the land use planners and purple for the ministry representatives.

@@

In order to integrate the results with the land use change model, we added a 5 km buffer around the linkages to simulate a corridor effect. This corridor width size is arbitrary and is in most cases  an unrealistic expectation for how large a wildlife corridor can or should be. \\

@@

The reforestation status modified what zones can be targeted for reforestation. In reforestation scenarios in Chapter 1, reforestation happened randomly (barring some neighboring rules). In this chapter, reforestation is targeted to the same areas identified in the workshop (all pixels within the identified priority areas and within a 5km buffer around the identified priority linkage areas). As in the model described in Chapter 1, all new reforested areas take the value of medium aged (30-50 yr) deciduous forest and remain in this state for the rest of the simulation. As previously mentioned, this assumption is necessary because LANDIS surface types are not available for newly forested areas. For the same reason, no forest transitions are allowed for this forest type. These two assumptions are conservative:  we do not impose succession or disturbance on these newly forested areas and assume medium age as an average age class.

@@

In order to be able to compare the results of these new scenarios with those developed in Chapter 1, we conducted the same connectivity analyses using the Circuitscape software. The software “borrows algorithms from electronic circuit theory to predict connectivity in heterogeneous landscapes". It is a free and open software under MIT license developed originally in Python, and now in its 5th version, in Julia \citep{circuitjulia}.

@@

Participants also workshopped constraints and assets to connectivity conservation in Montérégie. We divided the results into elements that had been spatialized (i.e. placed on a specific location of the map during the exercise), and non spatialized (i.e. not placed on a specific location), and scored them for the level of consensus they generated. Constraints scored higher, on average, than assets, indicating a focus of participants on constraints. The elements in the spatialized set scored higher on average than in the non spatialized set, indicating that participants tended to focus on more tangible and local issues and assets.

@@

Participants debated the importance of both "facilitating" and "constraining" factors in deciding whether a link is a priority. Facilitating factors included the feasibility of establishing the link, social facilitators such as political and social buy-in, and territorial facilitators, such as appropriate land use. There were several types of constraints: financial, social, territorial and physical, and must be considered in relation to the socio-political and spatial planning context in which they emerge.

@@

% Linear
\begin{figure}[h!]
\makebox[\textwidth]{
  \includegraphics[width=1.3\textwidth]{thesis/figures/connectivity_decrease_x5species_chap2.png}
}
 \caption{Change in mean flow (in \% of the 2010 flow) between 2010 (observed) and 2100 (predicted), contrasting BAU scenario (solid line) with other conservation scenarios - linear graph.}
 \label{fig:flow_linear_2}
\end{figure}

% Radar
\begin{figure}[h!]
\makebox[\textwidth]{
  \includegraphics[width=1.3\textwidth]{thesis/figures/radar_ggradar_chap2.png}
}
 \caption{Change in mean flow (in \% of the 2010 flow) between 2010 (observed) and 2100 (predicted), contrasting BAU scenario (solid line) with other conservation scenarios - radar graph.}
 \label{fig:flow_radar_2}
\end{figure}

@@

% Linear
\begin{figure}[h!]
\makebox[\textwidth]{
  \includegraphics[width=1.3\textwidth]{thesis/figures/surf_chap2.png}
}
 \caption{Change in the number of features (in \% of the 2010 flow) identified by the SURF analysis between 2010 (observed) and 2100 (predicted), contrasting BAU scenario with other conservation scenarios - linear graph.}
 \label{fig:surf_linear_2}
\end{figure}

% Radar
\begin{figure}[h!]
\makebox[\textwidth]{
  \includegraphics[width=1.3\textwidth]{thesis/figures/surf_radar_chap2.png}
}
 \caption{Change in the number of features (in \% of the 2010 flow) identified by the SURF analysis between 2010 (observed) and 2100 (predicted), contrasting BAU scenario with other conservation scenarios - radar graph.}
 \label{fig:surf_radar_2}
\end{figure}

### 2020-08-31 thesis/thesis.tex
small fixes

\documentclass[12pt,Bold,TexShade]{thesis/mcgilletdclass}
\usepackage{graphicx}
\usepackage{times}
\usepackage{subfig}
\usepackage{longtable}
\usepackage{float}
\usepackage{setspace}
\usepackage{multirow} % allows multiple rows in table
\usepackage{adjustbox}
\usepackage{enumitem}
\usepackage{rotating}
\usepackage{lipsum,lineno} % for placeholder text
\usepackage{array} % for sizing of tables
\usepackage{lscape} % for landscape tables
\usepackage{pdfpages} % to add pdf appendix
\usepackage{siunitx} % for km2 nottation

@@

\usepackage{hyperref} % for hyperlink accross the document

@@

This thesis therefore makes use of a hybrid modeling approach. We combine this model with scenarios which were derived from participatory methods. Next we introduce those methods and explain this integration. We end this introduction with simple predictions as to the results of our modelling effort. \\ \vspace{2em}

@@

Depending on the scenario, the model generates different forest change dynamics. The Historic scenario assumes no change in forest composition and stands as our control. While the baseline scenario assumes no further warming and status quo forest change dynamics, the RCP 8.5 scenario is one of the most severe climate change scenarios, and the trends observed in the baseline scenario are exacerbated \cite{ipcc_summary_2013}. We also derive two simple land use change scenarios: Business as usual (BAU) and Reforestation. Those two scenarios are meant to represent two extremes in land use change, and it is likely that the most probable trajectories for our landscape will be within those two boundaries. In combination with the climate scenarios, this brings our total scenario number to 6 (for chapter 1, as conservation scenrios are added in chapter 2). \\

@@

I now address some general limitations that concern both chapters. We argued that using markov-chain monte carlo (MCMC) based modelling for land use change allowed us to take into account the uncertainty of land use change simulations. The principle is to run the model multiple times (iterations), resulting in slightly different simulations every time. This allowed us to see that our model was quite constrained and that over 10 iterations we produced narrow confidence intervals when analysing the resulting connectivity; this was true for both chapters. However, there remains a large amount of uncertainty unaccounted for in the model. STSIM required a large number of parameters (for example, neighboring parameters, see \ref{tab:neigh_rules}) that were chosen and fixed to a certain value. To correctly account for uncertainty in those parameters, we could decide on a distribution for those parameters and draw from this distribution at each iteration. This is possible in STSIM, which allows the user to specify distributions. However this brings its own set of drawbacks. First, in order to get a representative sample from this distribution, more iterations would need to be computed, which would increase computing time. Second, the shape of the distribution for parameters like neighboring rules is not easy to determine a priori. This last point could potentially be alleviated by sampling the distribution from historical land use and land cover change data ; however, the question of how to conduct this sampling appropriately, for instance to account for spatial autocorrelation, would require more research. \\

@@

However, connectivity research can be hopeful in its ability to move beyond those limitations. Resistance-based models have been validated with genetic data \citep{marrotte_landscape_2014} and occurrence data (\cite{gray_landscape_2016}) and movement data \citep{lapoint_animal_2013}. Although genetic data alone cannot be expected to predict demographic connectivity without information on local demographic rates \citep{lowe_what_2010}, it can provide important information on connectivity when combined with other data sources. Connectivity modelling therefore needs to move toward new, integrative, and data-rich approaches. One example is combining multi-scale resistance surfaces from multiple data types, as \cite{zeller_multi-level_2017} did for pumas (\textit{Puma concolor}) in southern California. The role of novel forms of modelling, such as Individual-BasedModels (IBMs, a form of Agents-Based Models) is another potential direction for connectivity research: they can add realism to resistance based and circuit theory models and bring attention to mechanisms overlooked by those models, like mortality \citep{day_individual-based_2020}. In all cases, these approaches are data-intensive and seem to point to a need for more data on animal movements. \\

@@

But, as we quoted from \cite{williamson_spatially_2018} above, this view tends to assume that stakeholders want to pursue rational and optimal designs in their approach to conservation. There is a tendency to take a single-faceted approach to the optimisation problem, focusing on the trade-off between food production and the support for biodiversity. An alternative way is through the lens of ecosystem services  \cite{bennett_changing_2017}. Ecosystem services refer to the diversity of benefits that ecosystems provide to, and the many ways in which they support, human systems. Specifically, the concept of “bright spots'' has potential for guiding the discussion beyond dichotomies and single faceted approaches, by focusing on multifunctionality in agroecosystems. In a landscape, “bright spots” are areas exceeding expectations for goals, such as agro-ecological multifunctionality, biodiversity, or connectivity \citep{frei_bright_2018}. They show promise as a tool  to identify levers of change, and guide stakeholder discussion. Indeed, an approach that focuses on identifying bright spots for the provision of those ecosystem services can be valuable to an interactive process of stakeholder engagement. They offer starting points to work with to design scenarios. They are concrete examples to reason around what bundles of ecosystem services are desirable (elements of SW) for the conservation of ecological connectivity and which types of governance is needed to work toward them (elements of IC). In our study region, this framework is already being tested: \cite{frei_bright_2018} identified bright, dark and average spots within Montérégie for both landscape multifunctionality and avian biodiversity. They showed important spatial properties of bright spots, for instance that they were “neither spatially congruent nor in conflict”.  Connectivity could be fully integrated with this approach is two ways: first, it would be interesting to see if current bright spots are congruent with connectivity, in order to asses the mismatch between connectivity and other bundles of ecosystem services. One example is that of bee diversity which was found to be higher in connected habitats in Montérégie \citep{martins_complementary_2018}. Second, areas that are prioritized as important for connectivity under models such as the ones we developed in chapter 1 could be used in the identification of bright spots. \\

\citeauthor{frei_bright_2018} also pointed out a range of challenges for the identification of bright spots, the development of multicriteria measures being one of them. Connectivity measures are already being developed with  multifunctionality in mind. For instance, \cite{rayfield_priorisation_2018}, assessed whether certain habitat networks in the Montérégie would be more resilient to future change using several multi-scale metrics of connectivity. Another example is \cite{fletcher_towards_2019} who developed an unifying model for connectivity that takes into account multiple types of mobility. Connectivity is therefore a field that fits in the multicriteria framework of bright spots, as it itself moves in this direction.  Bright spots show promise for being integrated in more complex connectivity conservation planning tools. Finally, bright spots are intuitive, being based on statistical deviation from a given distribution and are therefore always relative to the context in which they are implemented. They are suited to the type of participatory framework we implemented in chapter 2: bundles of ecosystem services that are important to stakeholders can be identified and bright spots can be discussed, all in the iterative participatory setting we detailed above. \\

>>>>

\documentclass[12pt,Bold,TexShade]{thesis/mcgilletdclass}
\usepackage{graphicx}
\usepackage{times}
\usepackage{subfig}
\usepackage{longtable}
\usepackage{float}
\usepackage{setspace}
\usepackage{multirow} % allows multiple rows in table
\usepackage{adjustbox}
\usepackage{enumitem}
\usepackage{rotating}
\usepackage{lipsum,lineno} % for placeholder text
\usepackage{array} % for sizing of tables
\usepackage{lscape} % for landscape tables
\usepackage{pdfpages} % to add pdf appendix
\usepackage{siunitx} % for km2 notation

@@

\usepackage{hyperref} % for hyperlink across the document

@@

This thesis therefore makes use of a hybrid modeling approach. We combine this model with scenarios which were derived from participatory methods. Next we introduce those methods and explain this integration. We end this introduction with simple predictions as to the results of our modelling effort. \\ \vspace{1em}

@@

Depending on the scenario, the model generates different forest change dynamics. The Historic scenario assumes no change in forest composition and stands as our control. While the baseline scenario assumes no further warming and status quo forest change dynamics, the RCP 8.5 scenario is one of the most severe climate change scenarios, and the trends observed in the baseline scenario are exacerbated \cite{ipcc_summary_2013}. We also derive two simple land use change scenarios: Business as usual (BAU) and Reforestation. Those two scenarios are meant to represent two extremes in land use change, and it is likely that the most probable trajectories for our landscape will be within those two boundaries. In combination with the climate scenarios, this brings our total scenario number to 6 (for chapter 1, as conservation scenarios are added in chapter 2). \\

@@

I now address some general limitations that concern both chapters. We argued that using markov-chain monte carlo (MCMC) based modelling for land use change allowed us to take into account the uncertainty of land use change simulations. The principle is to run the model multiple times (iterations), resulting in slightly different simulations every time. This allowed us to see that our model was quite constrained and that over 10 iterations we produced narrow confidence intervals when analysing the resulting connectivity; this was true for both chapters. However, there remains a large amount of uncertainty unaccounted for in the model. STSIM required a large number of parameters (for example, neighboring parameters, see \ref{tab:neigh_rules}) that were chosen and fixed to a certain value. To correctly account for uncertainty in those parameters, we could decide on a distribution for those parameters and draw from this distribution at each iteration. This is possible in STSIM, which allows the user to specify distributions. However this brings its own set of drawbacks. First, in order to get a representative sample from this distribution, more iterations would need to be computed, which would increase computing time. Second, the shape of the distribution for parameters like neighboring rules is not easy to determine a priori. This last point could potentially be alleviated by sampling the distribution from historical land use and land cover change data ; however, the question of how to conduct this sampling appropriately, for instance to account for spatial autocorrelation, is not straightforward. \\

@@

However, connectivity research can be hopeful in its ability to move beyond those limitations. Resistance-based models have been validated with genetic data \citep{marrotte_landscape_2014} and occurrence data (\cite{gray_landscape_2016}) and movement data \citep{lapoint_animal_2013}. Although genetic data alone cannot be expected to predict demographic connectivity without information on local demographic rates \citep{lowe_what_2010}, it can provide important information on connectivity when combined with other data sources. Connectivity modelling therefore needs to move toward new, integrative, and data-rich approaches. One example is combining multi-scale resistance surfaces from multiple data types, as \cite{zeller_multi-level_2017} did for pumas (\textit{Puma concolor}) in southern California. The role of novel forms of modelling, such as Individual-Based Models (IBMs, a form of Agents-Based Models) is another potential direction for connectivity research: they can add realism to resistance based and circuit theory models and bring attention to mechanisms overlooked by those models, like mortality \citep{day_individual-based_2020}. In all cases, these approaches are data-intensive and seem to point to a need for more data on animal movements. \\

@@

But, as we quoted from \cite{williamson_spatially_2018} above, this view tends to assume that stakeholders want to pursue rational and optimal designs in their approach to conservation. There is a tendency to take a single-faceted approach to the optimisation problem, focusing on the trade-off between food production and the support for biodiversity. An alternative way is through the lens of ecosystem services  \cite{bennett_changing_2017}. Ecosystem services refer to the diversity of benefits that ecosystems provide to, and the many ways in which they support, human systems. Specifically, the concept of “bright spots" has potential for guiding the discussion beyond dichotomies and single faceted approaches, by focusing on multifunctionality in agroecosystems. In a landscape, “bright spots” are areas exceeding expectations for goals, such as agro-ecological multifunctionality, biodiversity, or connectivity \citep{frei_bright_2018}. They show promise as a tool to identify levers of change, and guide stakeholder discussion. Indeed, an approach that focuses on identifying bright spots for the provision of those ecosystem services can be valuable to an interactive process of stakeholder engagement. They offer starting points to work with to design scenarios. They are concrete examples to reason around what bundles of ecosystem services are desirable (elements of SW) for the conservation of ecological connectivity and which types of governance is needed to work toward them (elements of IC). In our study region, this framework is already being tested: \cite{frei_bright_2018} identified bright, dark and average spots within Montérégie for both landscape multifunctionality and avian biodiversity. They showed important spatial properties of bright spots, for instance that they were “neither spatially congruent nor in conflict”. Connectivity could be fully integrated with this approach in two ways: first, it would be interesting to see if current bright spots are congruent with connectivity, in order to assess the mismatch between connectivity and other bundles of ecosystem services. One example is that of bee diversity which was found to be higher in connected habitats in Montérégie \citep{martins_complementary_2018}. Second, areas that are prioritized as important for connectivity under models such as the ones we developed in chapter 1 could be used in the identification of bright spots. \\

\citeauthor{frei_bright_2018} also pointed out a range of challenges for the identification of bright spots, the development of multicriteria measures being one of them. Connectivity measures are already being developed with  multifunctionality in mind. For instance, \cite{rayfield_priorisation_2018}, assessed whether certain habitat networks in the Montérégie would be more resilient to future change using several multi-scale metrics of connectivity. Another example is \cite{fletcher_towards_2019} who developed an unifying model for connectivity that takes into account multiple aspects of animal mobility. Connectivity is therefore a field that fits in the multicriteria framework of bright spots, as it itself moves in this direction.  Bright spots show promise for being integrated in more complex connectivity conservation planning tools. Finally, bright spots are intuitive, being based on statistical deviation from a given distribution and are therefore always relative to the context in which they are implemented. They are suited to the type of participatory framework we implemented in chapter 2: bundles of ecosystem services that are important to stakeholders can be identified and bright spots can be discussed, all in the iterative participatory setting we detailed above. \\
